Malware Detection Via Machine Learning

### 2017-06-26 project_report/main.tex
Work at project submission

\documentclass{llncs}
\usepackage[utf8]{inputenc}
\usepackage{todonotes}
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{dirtytalk}
\usepackage{color}
\usepackage{glossaries}

@@

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%  HEADER  %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\title{Malware Detection Via Machine Learning}
\subtitle{Some Clever Subtitle?}
\titlerunning{Malware Detection Via ML}
\author{João F. Godinho\inst{1} \and Pedro Adão\inst{2}}
\institute{Instituto Superior Técnico\\
\email{enter email}\and
Instituto Superior Técnico\\
\email{enter email}}
\maketitle
\todo[inline]{Validate the title, subtitle, authors, email}

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%% ACRONYMS %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\newacronym{caro}{CARO}{Computer Antivirus Research Organization}
\newacronym{ai}{AI}{Artificial Intelligence}
\newacronym{ml}{ML}{Machine Learning}

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%% ABSTRACT %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
The abstract should summarize the contents of the paper
using at least 70 and at most 150 words.
\end{abstract}

\clearpage
\todo[inline]{Table of contents?}

\mainmatter
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%  INTRO   %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{sec:introduction}
% #1 Semantics-Aware Malware Detection
% #2 Attacking Malicious Code: A report to the Infosec Research Council
% #3 Malware, Rootkits & Botnets A Beginner's Guide pp.10
% #4 Computer viruses: theory and experiments
% #5 Theory of self-reproducing automata
% #6 The Evolution of Viruses and Worms
% #7 The art of computer virus research and defense
% #8 https://www.av-test.org/en/
% #9 http://www.caro.org/articles/naming.html
% #10 http://members.chello.at/erikajo/vnc99b2.txt
Malware, which can be referred to by other names like malicious software or malicious code, is not easily described by a single definition. For example, Christodorescu et al.[quote 1] simply describe malware as \say{a program that has malicious intent}, which in turn begs the question, what is malicious intent? McGraw and Morrisett[quote 2] give a more concrete definition by saying \say{Malicious code is any code added, changed, or removed from a software system in order to
intentionally cause harm or subvert the intended function of the system}. Both definitions give an overall understanding of \textcolor{red}{(on?)} what is malware, but to better understand the definition, one can briefly look at the history of malware.

\todo[inline]{Clean previous paragraph. Remove bellow}

The term malware was first used by Yisrael Radai in 1990[quote 3], before then malicious software was referred to as computer viruses, a notion which was first formalized by Cohen in 1983[quote 4]. Before Cohen, John Von Neumann had already done some similar academic work that closely relates to computer viruses, namely \say{Theory of self-reproducing automata}, published in 1699[quote 5]. The term computer virus is one of the many types of malicious software, but given that it predates the term malware, it is not uncommon to see both terms used as interchangeably.

Even before the formalization of computer virus by Cohen in 1983, one of the first documented computer virus, although experimental, was developed by Bob Thomas at BBN in 1971[quote 6]. The program was named Creeper and consisted of a self-replicating program that would print a message and move between the nodes of ARPANET.

Another malicious software worth mentioning, given it was one of the first known virus that spread \say{in the wild} (i.e. outside the computer system or laboratory in which it was developed), is Elk Cloner. Created by Rich Skrenta in 1982[quote 7], Elk Cloner spread by floppy disk and had a payload that displayed Skrenta's poem after every 50th use of the infected system.

\todo[inline]{Up to here}

Since then, the amount of malware has grown at an alarming rate. According to AV-TEST[quote 8], an independent security software group, up to 2012 the amount of total malware per year would not surpass 100 million, whereas 2013 almost reached 200 million, followed by more than 100 million increase per year, with almost 600 million total malware before the end of 2016.

\todo[inline]{Clean previous paragraph. Remove bellow}

Even though the term malware generalizes the malicious intent of a piece of software, one can classify malware further based on its purpose and behavior. Given the high amount of malware, there is also a high variety of malware, based on purpose and behavior.

From the general purpose and behavior of a malware sample, it can be classified as belonging to one or more malware categories (or classes). Examples of categories are virus, worms, trojans and adware, to name a few. Having the ability to classify malware into a more restrict subset is helpful, but there is still a need to distinguish between samples of the same category, by giving more detailed names to malware.

One attempt at creating a naming convention was made in 1991, by the Computer Antivirus Research Organization (CARO). The convention states the name should consist of four parts, separated by a dot. The four constituents are the family name, group name, major variant name and minor variant name, displayed in that order. An optional modifier could also be appended to the name with a colon. The problem with the aforementioned convention is that it only covered computer viruses, which were more prominent at the time.

An attempt at improving CARO's naming convention was made in 1999, by Gerald Scheidl[quote 10]. His proposal added two other constituents, platform and type, as a prefix to CARO's naming.

Later attempts to standardize malware naming were made, none of which resulted in a standard widely adopted. Due to this, antivirus vendors use their own naming scheme, which are variants of the mentioned schemes.

Since the malware naming scheme is not standardized, and given the amount and variety of malware, different family names with the same behavior emerge from different antivirus vendors. The name redundancy is worsen by how different vendors detect and classify malware.

With the raise of malware amount and diversity, being able to identify and classify malware becomes essential in order to protect systems.

Before the late eighties, given the low amount of malware, the first antivirus programs were designed to find and remove a handful of malware, mainly through signature based methods, which detect malware by looking for specific patterns that appear in known malware.

In the late eighties and nineties the first antivirus were released, along with more sophisticated detection methods that make use of heuristics, which also detect malware by also looking for patterns, but at a higher level of abstraction.

In more recent years, and given the increment of malware amount and variety, signature and heuristic methods fall short when keeping up with new types of malware. The mentioned methods work best for samples of known malware or new malware with small variants of known malware. This is because these methods look for known patterns. If a new type emerges, which pattern's are not known, it is not flagged as malicious.

Given this limitation, more advanced methods like behavioral analysis are applied together with the previous mentioned methods. Behavioral analysis takes into account the actions performed by the malware, introducing dynamic information, from which patterns can be extracted and again matched against known patterns.

Having to run samples in order to extract dynamic information poses the problem of infecting the system running the samples. One way to prevent this is to use a sandbox environment, which is a security mechanism to isolate running programs. Using a sandbox environment closely relates to computer virtualization.

To provide users the possibility to check files without running the risk of being infected, online services allow users to submit files which are then run in a sandbox environment.

An example of such service is malwr. Malwr takes user submissions and runs them in a sandbox environment, specifically Cuckoo Sandbox, providing static and dynamic information about the sample, as well as checking if the file is known in VirusTotal, a similar service that also checks if the submission is known in different antivirus solutions.

% VirusScan (from McAfee) and NOD (from ESET) were first released in 1987; AntiVir (from Avira) and Avast Antivirus were released in 1988. Norton Anti-Virus (from Symantec) and F-Secure antivirus were released in 1991.

\todo[inline]{re-focus on high amount of malware, difficulty of having to run all samples and check}
\todo[inline]{propose machine learning as a solution to deal with the high amounts of data}
\hfill \break

\todo[inline]{Motivation}
\begin{itemize}
	\item Define Malware, problem with that, subjectivity
	\item Quantify malware in recent years (quote vendors/get stats from header analysis)
	\item What is being done about malware?
	\item How is malware identified?
	\begin{itemize}
		\item Types of strategies
		\item Is it effective? efficient?
	\end{itemize}
	\item Being able to identify malware as early as possible would help?
	\item Introduce services like malwr.com/virustotal.com
	\item Introduce Machine Learning capabilites
	\item Automation of detection/classification
\end{itemize}

%From an external view, adware and ransomware will provide a noisy output to the user, given their function to deliver unwanted advertisements (for adware) and request data ransom (for ransomware). Spyware and rootkit types, on the other hand, will provide little to no user interaction, given their function to spy on user activity (for spyware) and conceal processes, files or system data (for rootkit).
%
%From an internal view, adware and ransomware will behave similarly, since both need interfaces that facilitate user interaction (e.g. create and show dialogs), in contrast to spyware and and rootkit, which thrive on concealing actions (e.g. run hidden processes).

% Quote this for more types of malware
% https://support.symantec.com/en_US/article.HOWTO54185.html

\todo[inline]{Problem}
\begin{itemize}
	\item Remove the human factor from detection/classification to speed up process.
	\item Introduce lack of temporal consistency here?
	\item Talk how related work creates own corpus and low corpus size/variation?
	\item Refer stats from header analysis? Time vendors take to take knowledge of samples/Time until new malware is detected in most vendors
	\item Reinforce time to classify malware?
\end{itemize}

\todo[inline]{Hypothesis}
\begin{itemize}
	\item Would a more automated approach, based on ML, facilitate malware detection/classification?
	\item Dynamic/Static analysis to detect and classify malware
	\item Temporal consistency is a factor that should be taken into account?

\end{itemize}
\todo[inline]{Objectives/Contributions}
\begin{itemize}
	\item Usage of real world analysis
	\item Try to apply ML to PE32 analysis from malwr.com (Talk about this here?)
	\item Contribute with a corpus of analysis from malwr.com
	\item Maintain a temporal consistency

\end{itemize}
\todo[inline]{Document outline}
The following sections are organized as follows. Section \ref{sec:related_work} describes related work and its influence on the current work. Section \ref{sec:solution} describes the planned approach for solving the problem at hand. Section \ref{sec:conclusion} closes the report and references the work done so far.

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%% RELATED  %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work}\label{sec:related_work}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Background
\subsection{Background}\label{subsec:background}
The term malware was first used by Yisrael Radai in 1990[quote 3], before then malicious software was referred to as computer viruses, a notion which was first formalized by Cohen in 1983[quote 4]. Before Cohen, John Von Neumann had already done some similar academic work that closely relates to computer viruses, namely \say{Theory of self-reproducing automata}, published in 1699[quote 5]. The term computer virus is one of the many types of malicious software, but given that it predates the term malware, it is not uncommon to see both terms used as interchangeably.

Even before the formalization of computer virus by Cohen in 1983, one of the first documented computer virus, although experimental, was developed by Bob Thomas at BBN in 1971[quote 6]. The program was named Creeper and consisted of a self-replicating program that would print a message and move between the nodes of ARPANET.

Another malicious software worth mentioning, given it was one of the first known virus that spread \say{in the wild} (i.e. outside the computer system or laboratory in which it was developed), is Elk Cloner. Created by Rich Skrenta in 1982[quote 7], Elk Cloner spread by floppy disk and had a payload that displayed Skrenta's poem after every 50th use of the infected system.

Over the years, malware has grown in size and diversity, which introduces the need to be able to distinguish malware, not only to facilitate its removal, but also to provide safer mechanisms of defense.

Based on the malware general purpose and behavior, a high level classification of the malware can be made by splitting malware into one or more categories (or classes). Examples of such classes are viruses, worms, trojans and adware, to name a few. This high level classification is helpful, but lacks detail, like target platform, origin or version. In order solve this issue, attempts have been made to create a naming convention for malware.

One of the first attempts was made in 1991, by the \gls{caro}\footnote{http://www.caro.org/articles/naming.html}, which states the name should consist of four  obligatory parts, separated by a dot, and an optional fifth part. The four main constituents are the family name, group name, major variant and minor variant name. The optional fifth part is a modifier and could be appended to the name with a colon.
\todo[inline]{Describe the parts?}

Although the aforementioned naming convention is well defined, it is aimed at naming viruses, lacking differentiation for other classes as well as target platform. In order to solve this, an improvement was proposed in 1999 by Gerald Scheidl\footnote{http://members.chello.at/erikajo/vnc99b2.txt} that suggested the addition of a prefix that consisted of the platform and type.

A more recent attempt at creating a naming convention was made during the Virus Bulletin Conference in 2001\footnote{https://www.virusbulletin.com/uploads/pdf/magazine/2002/200201.pdf}, with the idea that malware naming should consist on three levels of information: malware name, classification and full text description.
\todo[inline]{Again, describe parts?}

None of the referenced naming conventions is widely adopted, with antivirus vendors having their own naming conventions, that although similar, vary in some way or another. For example, Trend Micro\footnote{http://docs.trendmicro.com/all/ent/smex/v11.0/en-us/smex\_11.0\_olh/secrisk\_malware\_name.html} and Symantec\footnote{https://www.symantec.com/security\_response/virusnaming.jsp} use a prefix (which identifies the malware class), threat name (i.e. family) and suffix (which indicates a variant of the threat). Whereas Microsoft\footnote{https://www.microsoft.com/security/portal/mmpc/shared/malwarenaming.aspx} tries to follow the \gls{caro}'s convention, by using a type, platform, family, variant and information. Avira\footnote{https://www.avira.com/en/support-malware-naming-conventions} on the other hand, takes a more simplistic approach by using a prefix (which identifies the malware type) and a threat name.

Although the naming and its constituents are similar between vendors, no convention is set, hence the threat names given to malware varies, increasing the redundancy for the same malware. According to an article published by Symantec in 2002\footnote{https://www.symantec.com/connect/articles/virus-any-other-name-virus-naming-practices}, based on a list of 300,000 unique names, the name redundancy is quantified as an average of four different names for the same malware.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Malware detection
\subsection{Malware Detection}\label{subsec:malware_detection}
In the early days of malware, due to low amount and variety, antivirus could target malware specifically (e.g. Reaper was created to remove Creeper), but with an estimate of more than 600 million total malware by the end of 2016, targeting specific malware becomes increasingly more difficult, hence being able to identify and classify malware efficiently is crucial to protect systems.

What follows is a quick rundown on different detection techniques developed over the years[quote 7].

\subsubsection{First-Generation Scanners.} These methods of detection work at a fairly simple level and use static methods of detection. The applied principles include:
\begin{itemize}
	\item String scanning, where a sequence of bytes that is common in a malicious program but not in a regular program is matched against the suspicious file.
	\item Wildcards, similar to string scanning, but instead of a fixed sequence of bytes, regular expressions can be used in the sequence of bytes, this allows variation and byte skipping when matching.
	\item Mismatches, also similar to string scanning, in which a sequence of bytes is taken as reference, but a number of bytes $N$ is given allowing up to $N$ bytes to vary in any position of the string (e.g. the string \say{\texttt{01 02}} with $N=1$ will match \say{\texttt{XX 02}} or \say{\texttt{01 XX}}).
\end{itemize}

The described principles can be applied with slight variations that improve the performance of the scanner, such as matching on the beginning and end of the file (top-and-tail scanning) or matching at a certain offset (bookmarks).

They all require a knowledge base with known patterns that must be kept up-to-date in order to detect newer malware.

\subsubsection{Second-Generation Scanners.} These methods improve on the previous generation by doing a static analysis at a higher level of abstraction. Where first-generation scanners would match sequence of bytes, second-generation scanners also look at the representation of the bytes. The applied principles include:
\begin{itemize}
	\item Smart scanning, where the Assembly representation of the bytes is taken into account and instructions that have no effect (e.g. \texttt{NOP}) can be skipped. The resulting signature is smaller and more easily detects mutations. The same principle could be applied to textual malware (e.g. scripts), where extra white spaces have no effect.
	\item Nearly exact identification, where instead of relying on a single string for detection, double-string detection can be used. An alternative to double-string detection could also be based on a checksum range selected from the malware body, giving a higher accuracy by taking into account a larger portion of the malicious file.
	\item Exact identification, which can exactly identify malware variants, is usually used in combination with first-generation methods and instead of using a single range like nearly exact, uses multiple ranges to calculate checksums. To be accurate, this method must only take into account constant bits of the malicious file, and given that more ranges are used it is slower then the previous ones.
\end{itemize}

Like in the previous generation, a knowledge base must be used and kept up-to-date to identify newer malware.

\subsubsection{Algorithmic Scanning Methods.} These methods are not so focused on direct matching of signatures, but on the routines implemented in the scanner that allow the detection of malware. Early implementations of algorithmic scanning were hard-coded in the antivirus, but with different types and families emerging, the hard-coded algorithms could easily become obsolete as such. To overcome this, vendors would introduce more specific routines for malicious programs when updating the antivirus.

Being able to update the antivirus algorithms frequently enables the vendors to have a faster response to new malware by developing detection routines for new malware and pushing them to end users.

\subsubsection{Code Emulation.} This detection technique uses dynamic analysis to detect malware by simulating code execution. A virtual machine is used to simulate the system's components, thus the malicious files are encapsulated and no code is executed by the real processor.

Code emulation can be faster than using algorithmic scanning methods when dealing with encrypted/obfuscated malware, as the code can be emulated and the decrypted/deobfuscated code is much easier obtained.

\subsubsection{Heuristic Analysis.} This detection technique takes into account several characteristics of malicious software, both static and dynamic. Examples of such characteristics are code execution starting in uncommon sections, suspicious code redirection or incorrect size of code in header.

When checking a suspicious file, the presence or absence of multiple characteristics is weighted to assert if the file is deemed malicious or not. The upside of this technique is that it allows for never seen malware to be correctly flagged, but brings the downside of a higher amount of false positives.

>>>>

\documentclass{llncs}
\usepackage[utf8]{inputenc}
\usepackage{todonotes}
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{dirtytalk}
\usepackage{color}
\usepackage{glossaries}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage[caption = false]{subfig}
\graphicspath{ {images/} }
\usepackage{rotating}
\usepackage{pdflscape}

@@

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%  HEADER  %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\title{Malware Detection Via Machine Learning}
\subtitle{}
\titlerunning{Malware Detection Via ML}
\author{João F. Godinho}
\institute{Instituto Superior Técnico\\
\email{joao.f.godinho@tecnico.ulisboa.pt}}
\let\oldaddcontentsline\addcontentsline
\def\addcontentsline#1#2#3{}
\maketitle
\def\addcontentsline#1#2#3{\oldaddcontentsline{#1}{#2}{#3}}
%\todo[inline]{Validate the title, subtitle}

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%% ACRONYMS %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\newacronym{hids}{HIDS}{Host-based Intrusion Detection Systems}
\newacronym{ml}{ML}{machine learning}
\newacronym{caro}{CARO}{Computer Antivirus Research Organization}
\newacronym{ai}{AI}{Artificial Intelligence}
\newacronym{api}{API}{Application Programming Interface}
\newacronym{bios}{BIOS}{Basic Input/Output System}
\newacronym{dll}{DLL}{Dynamic-Link Library}
\newacronym{cc}{CC}{Creative Common's}
\newacronym{ann}{ANN}{Artificial Neural Networks}
\newacronym{tpr}{TPR}{True Positive Rate}
\newacronym{fpr}{FPR}{False Positive Rate}
\newacronym{svm}{SVM}{Support Vector Machine}
\newacronym{mist}{MIST}{Malware Instruction Set}
\newacronym{lr}{LR}{Logistic Regression}
\newacronym{pe}{PE}{Portable Executable}

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%% ABSTRACT %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
	Malware, which focuses on harming or subverting a system, has grown exponentially in recent years. The regular signature and heuristic methods applied by anti-virus vendors can fall short on keeping up with the rise of new types of malware. Other works have shown that machine learning has potential for dealing with the task of malware detection. With this in mind, the current report elaborates on the problem of malware, how anti-virus detect it, and proposes the development of a machine learning system to deal with the task of malware detection. To design the system to be, the current work will make use of a dataset with 388,513 samples for training and evaluation, focusing on how laboratory and real-world conditions can affect the final result.
\end{abstract}

\setcounter{tocdepth}{2}
\tableofcontents

\mainmatter
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%  INTRO   %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%

% Temporary quotes, move to bib
% 1 - Attacking Malicious Code: A report to the Infosec Research Council
% 2 - AV-TEST Security Report 2015/2016 - https://www.av-test.org/fileadmin/pdf/security_report/AV-TEST_Security_Report_2015-2016.pdf
% 3 - McAfee Labs Threats Report April 2017 - https://www.mcafee.com/us/resources/reports/rp-quarterly-threats-mar-2017.pdf
% 4 - Symantec Internet Security Threat Report - https://www.symantec.com/content/dam/symantec/docs/reports/istr-22-2017-en.pdf
% 5 - Scanners of The Year 2000: Heuristics - http://vxheaven.org/lib/adg00.html
% 6 - Thomas Mitchell - Machine Learning
% 7 - Some Studies in Machine Learning Using the Game of Checkers - Arthur L. Samuel - http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.368.2254&rep=rep1&type=pdf
% 8 - Prudent Practices for Designing Malware Experiments: Status Quo and Outlook

\section{Introduction}\label{sec:introduction}
Malware, which can be referred to by other names like malicious software or malicious code, can be described as \say{any code added, changed, or removed from a software system in order to
intentionally cause harm or subvert the intended function of the system}\cite{mcgraw:mal_code}. Malware can be a piece of code attached to a legitimate program, an independent program, or a combination of both. The reasoning that leads to malware creation varies greatly, some are created to demonstrate some vulnerability or concept and do not cause direct harm to systems, while others are used to steal information or render systems useless. The wide variety and quantity of now-a-day systems and the amount of sensitive information they store provide numerous opportunities to illegally make a profit out of subverting legitimate systems.

The amount of malware has shown an exponential increase since the early 2000s, from 1.7 million in 2005 to over 600 million by the end of 2016\cite{av-test:report,mcafee:report}. According to a 2007 study by McAfee\footnote{McAfee Study Finds 4 Percent of Search Results Malicious, by Frederick Lane, June 4, 2007 [http:\/\/www.newsfactor.com\/story.xhtml?story\_id=010000CEUEQO]} about 4\% of the query results by major search engines lead to potentially dangerous websites. A more recent study by Norton in 2010\footnote{Symantec Press Release, July 29, 2010 [https://www.symantec.com/en/aa/about/newsroom/press-releases/2010/symantec\_0729\_02]} showed that at least 10\% of top trending search terms returned malicious results. From the contact with malicious software, 1.1 billion identities (i.e. user's personal information) have been exposed in 2016 alone\cite{symantec:report}.

Given the wide range and quantity of systems being operated by naive users, in the sense that they lack the knowledge to distinguish malicious and legitimate software, the first line of defense against threats are \gls{hids}, commonly known as anti-virus solutions. Their main function is to prevent infections by malicious code, which is achieved by detecting a program as malicious and disallow it from running (i.e. quarantining the malicious file).

To detect malicious software, anti-virus solutions depend on two main methods. \textit{Signature-based} methods check for the presence of unique byte sequences from known malware. For this method to be effective, a database of all malware signatures must be available, which leads to an obvious shortcoming: only known malware is detected\cite{gryaznov:scanners}. New and unknown malware is not detected until vendors analyze it and update their clients. To improve the effectiveness of anti-virus and overcome the problem in aforementioned methods, \textit{heuristic-based} methods are employed. These apply a set of rules to decide whether a program is likely to be malware or not. Like before, a database of rules must be available, but unlike signature methods, which provide exact matching on malware, heuristic methods are capable, to some degree, of generalizing malware features, hence detecting new and unknown malware. Although more sophisticated, heuristic-based methods need a database of rules that must be defined \textit{a priori}, just like signature methods.

As mentioned before, signature methods only detect known malware, which translates to a very low false positive rate (i.e. goodware wrongly classified as malware), while having a high detection rate (i.e. correctly classified malware from the total malware), assuming the signature database is on par with malware samples, which at the rate new malware emerges becomes a hard task. For heuristic methods new and unknown malware can be detected due to similarities to known malware, translating to a higher detection rate with less information, but with the downside of raising the false positive rate, as benign programs (goodware) can be flagged malicious due to characteristics that fit into malware rules\cite{gryaznov:scanners}. To improve the effectiveness of anti-virus, both methods must complement each other.

Having the need to either create signatures for new malware or define a set of rules that describe malware characteristics becomes a cumbersome task for the increasing malware numbers. To try and improve the detection task of anti-virus software, the current work proposes a \gls{ml} approach to malware detection as an evolution to heuristic-based methods.

The machine learning field studies the development of computer programs that automatically improve with experience. A program is given a task and an experience, if its performance at the task improves with the experience given, then the program is learning the task\cite{mitchell:ml}. In the proposed work, the task at hand is to distinguish malware from goodware (i.e. classify samples) and the experience is a wide variety of samples, both goodware and malware. The final result is a program that performs well for the given task (i.e. successfully learns to distinguish goodware from malware).

The \gls{ml} field has been around for multiple decades, with some of the earlier works dating back to 1959 by Arthur L. Samuel\cite{samuel:studies}, but only in recent years has the field grown in popularity. Its recent growth can be accredited to two main factors: availability of large data sets and computation power. Larger data sets provide more experience for a program to learn from, but also require more processing power.

The growing number of malware samples provide a good opportunity to make use of \gls{ml} for malware detection. This approach is not new and has provided high detection rates with low false positive rates\cite{miller:rev_int,nissim:al_pdf,perdisci:behavior,rieck:dynamic,schultz:data_mining}. However, many of the resulting research is done under laboratory conditions that do not represent real world scenarios. Some of those conditions have been studied in \cite{rossow:practices} and include the correctness of datasets, transparency and realism. Alongside the application of \gls{ml} to detect malware, the proposed work will study how laboratory conditions differ from real world one's and how effectiveness is affected.

To provide a solution that makes use of \gls{ml} to detect malware and its performance under laboratory \textit{vs.} real world conditions, the proposed work will follow an approach that is divided into three stages:

\begin{itemize}
	\item \textit{Data collection}: experience from which a program can learn must be available, as such the first step must ensure the collection of samples that represent both malware and goodware examples. The current work will take advantage of the Malwr\cite{tool:malwr} service, which provides both static (e.g. library imports) and dynamic (e.g. library call order during runtime) details about a sample. The total number of available samples surpasses 388 thousand, dated from April 2013 to October 2016.
	\item \textit{Data labeling}: before training a classifier to detect malware, samples must be labeled as either benign or malign. The Malwr\cite{tool:malwr} service provides details on how multiple anti-virus solutions classify a sample, hence the approach is to use said classification to determine the legitimacy of a sample. This step is not linear, as different anti-virus classify samples differently, introducing edge cases (e.g. 50\% classify malware, the other 50\% classify goodware). To overcome this, the current work must develop a metric to deal with such cases.
	\item \textit{Classifier implementation}: at this stage the collected data is ready to be used as experience for the program. Using static and dynamic details together with the labeling of the samples, a classifier can be trained to distinguish between malware and goodware. Alongside the classifier creation, this phase must also measure the effectiveness of the solution. For this work's scope, the effectiveness is to be measured under laboratory and real world conditions.
\end{itemize}

In sum, the work to be done as to verify the application of \gls{ml} methods for the task of detecting malware is: collection of samples, labeling of collected samples and finally usage of the resulting dataset to achieve a malware classifier.

The following sections provide deeper analysis on the subject of using \gls{ml} to classify malware, as well as the choices to be made and the reasoning behind them. Section \ref{sec:related_work} describes related work that both influenced and motivated the current work. Section \ref{sec:solution} details the intended approach, the tools to be used, how the work is to be presented and the already accomplished work. Section \ref{sec:conclusion} summarizes and closes the report.

%\todo[inline]{falar dos antivirus para combater isto, falar nas tecnicas basicas: signature based, heuristic based. tirar partido de ML para estender metodos heuristicos e melhorar a classificaçao de malware.}
%
%
%\todo[inline]{one liner do que é malware (quote), quantificar (quote), necessidade de detectar malware mais rapidamente (automatizar/generalizar).
%
%Avanços na detecçao do malware, pouca coisa. Aplicaçao de ML neste tipo de problema.
%
%Contributos do trabalho para este problema, desenvolver um classificador que funcione no mundo real e ajude a resolver o problema.
%
%1. Dados recolhidos, o que posso extrair dos dados\\
%2. Definir o que é malware (thresholds, consenso)\\
%3. Comparaçao de metodologias (estatatica/dinamica)}

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%% RELATED  %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work}\label{sec:related_work}

% #1 Semantics-Aware Malware Detection
% #2 Attacking Malicious Code: A report to the Infosec Research Council
% #3 Malware, Rootkits & Botnets A Beginner's Guide pp.10
% #4 Computer viruses: theory and experiments
% #5 Theory of self-reproducing automata
% #6 The Evolution of Viruses and Worms
% #7 The art of computer virus research and defense
% #8 https://www.av-test.org/en/
% #9 http://www.caro.org/articles/naming.html
% #10 http://members.chello.at/erikajo/vnc99b2.txt
% #11 Learning to Detect and Classify Malicious Executables in the Wild
% #12 The Evolution of Viruses and Worms
% #13 http://cert.europa.eu/static/SecurityAdvisories/2017/CERT-EU-SA2017-012.pdf
% #14 https://www.virusbulletin.com/uploads/pdf/magazine/2002/200201.pdf

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Background
\subsection{Background}\label{subsec:background}
Malware, which can be referred to by other names like malicious software or malicious code, can be described as \say{a program that has malicious intent}\cite{christodorescu:semantics}, or more detailed as any code added, changed, or removed from a software system in order to intentionally cause harm or subvert the intended function of the system\cite{mcgraw:mal_code}.

The term malware was first used by Yisrael Radai in 1990\cite{elisan:malware}, before then malicious software was referred to as computer viruses, a notion which was first formalized by Cohen in 1983\cite{cohen:virus}. Given term computer virus predates the term malware, it is not uncommon to see both terms used interchangeably. For the purpose of this work the term malware is adopted, for reasons explained in the following paragraphs.

To the exponential growth of malware\cite{av-test:report}, the ability to distinguish between malware samples is crucial. In the biological world, different types of infections reckon distinctive disinfections, the same applies to the digital world. The infection method, behavior and subsequently purpose of malware samples vary, hence being able to make a distinction facilitates prevention and disinfection by anti-virus solutions.

There exists no ground truth when it comes to distinguishing malware, leading to subtle differences on the classification and naming of the same malware instance by different parties. To facilitate the reader's understanding of the topic, the current work will focus on defining malware types based on their propagation and purpose.

With respect to the propagation method, malware can be split into three classes\cite{kolter:learning}.

\paragraph{Virus.}\footnote{This work uses virus as a subset of malicious code, hence the choice of using the term malware over virus.}This type of malware inherits its name from resembling to biological viruses. A virus is usually composed by two main subroutines. The first subroutine is responsible for infecting other programs by attaching the virus code, while the second subroutine is the actual malware payload that contains the malware purpose (i.e. virus payload)\cite{chen:evolution}. Viruses are attached to programs and propagate when an infected program is run, hence depending on other programs (as hosts) and user interaction. For viruses to infect new systems, an infected file must be carried between them by a user (e.g. USB pen, email attachment), reinforcing the user's role in this propagation method.

\paragraph{Worm.} The increasing number of network connected devices facilitates the application of worms to propagate malware. A worm shares the self-replicating ability of viruses, but discards user interaction and the need for a host. This is possible by making a worm an independent program that exploits networks to find vulnerable systems to infect with a copy of themselves\cite{chen:evolution}. Similarly to viruses, worms can contain additional payload to perform malicious actions on infected systems (e.g. Blaster worm exploited a vulnerability in RPC for propagation while its payload flooded a Microsoft domain\footnote{W32/Blaster worm, CERT, August 11, 2003 [http://www.cert.org/historical/advisories/CA-2003-20.cfm]}).

\paragraph{Trojan.} While viruses and worms focus on self-propagation, trojans focus on deceiving users into executing them, disregarding propagation. Trojans try to appeal users with some useful functionality as to allure into running the program\cite{szor:art}. By hoaxing users into running them, trojans bypass the need for a host (in the virus case) or an exploit (in the worm case) to perform malicious actions.\\

These three classes are not exclusive and can be used in conjunction to facilitate propagation (e.g. a virus that relies on a trojan to start propagating). Given the inexistent standards for malware naming, some literature\cite{szor:art} define worms as a subset of viruses, as some worms start by attaching to files. With the general infection methods layered out, the following will succinctly describe the general naming when referring to the malware purpose (i.e. its payload).

Given the high numbers of malware multiple purposes imply multiple names, as such the following lists some of the most commonly heard names and definitions as given by McAfee\cite{mcafee:glossary}:

\begin{itemize}
	\item \textit{Adware}: Program that automatically displays advertisement to the user. Most do not cause direct harm.
	\item \textit{Backdoor}: Installs or takes advantage of an unknown entry point to a user's system, allowing remote control of a system.
	\item \textit{Dropper}: Program that facilitates the introduction of other malware instances into a system.
	\item \textit{Keylogger}: Program that monitors a user's input (e.g. keyboard strokes) to steal private information.
	\item \textit{Password Stealer (PWS)}: Program that specifically targets a user's personal information, like usernames and passwords.
	\item \textit{Ransomware}: Program that restricts access to a system/files (usually by encryption) and demands a ransom to allow a user his access.
	\item \textit{Rootkit}: Program designed to hide the existence of other (malicious) programs, hindering detection of malicious applications.
	\item \textit{Scareware}: Program that fakes the existence of malware and tricks the user into paying for disinfection or into downloading real malware.
	\item \textit{Spyware}: Program that spies on a user's activity through multiple vectors (e.g. taking prints of a user's desktop/webcam).
\end{itemize}

The ability to name a malware given its propagation method and purpose facilitates detection and disinfection, but given the high quantities of malware, being able to particularly discriminate an instance is helpful. To that purpose, attempts have been made to create naming conventions for malware.

One of the first attempts was made in 1991, by the \gls{caro}\cite{caro:naming}. This attempt was aimed at viruses only and focused on discriminating and relating virus samples. The convention defines an hierarchical naming with three levels. The base of the hierarchy is the family name, where structurally similar virus should be grouped. To further distinguish instances, a group name and major variant name is applied. To differentiate versions of the same instance and provide additional information about a sample, minor variant name and modifier fields are to be used. In sum, a virus name should look as following:

\begin{center}\texttt{family.group.major.minor:modifier}\end{center}

To enable differentiation from other types of malware, an improvement to \gls{caro}'s naming was proposed in 1999 by Geral Scheidl\cite{scheidl:naming}. Scheidl's proposal included a prefix that consisted of the platform targeted by the malware and a type (i.e. previously defined classes and names), rendering a more in-depth name:

\begin{center}\texttt{platform.type/family.group.major.minor:modifier}\end{center}
